---
title: "Analysis of the Adverse Health and Economic Impacts of US Storms (NOAA Storm Database)"
author: "Murali"
date: "2/7/2020"
output: 
  html_document:
    keep_md: true
    self_contained: true
---
##global options
```{r}
knitr::opts_chunk$set( fig.path='figure/', echo = TRUE, cache = TRUE )
```

## Introduction

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

## Synopsis
The goal of the assignment is to explore the NOAA Storm Database and explore the effects of severe weather events on both population and economy.The database covers the time period between 1950 and November 2011.

The following analysis investigates which types of severe weather events are most harmful on:

1. Health (injuries and fatalities) 
2. Property and crops (economic consequences)

Information on the Data: [Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

## Data Processing

### Loading Data
Download the file and extract the data into a dataframe
```{r}
library("data.table")
library("ggplot2")

fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
destinationFiile = "repdata%2Fdata%2FStormData.csv.bz2"
download.file(fileUrl, destfile = destinationFiile)
stormCSV <- read.csv("repdata%2Fdata%2FStormData.csv.bz2")

# Converting dataframe to data.table
stormDataSet <- as.data.table(stormCSV)
```

### Exploring Dataset
Take a look at the column names
```{r}
colnames(stormDataSet)
```
### Subsetting
Create a dataset with the columns that we are interested
```{r}
stormDataSet <- stormDataSet[ ,c("EVTYPE", "FATALITIES", "INJURIES"  , "PROPDMG" ,   "PROPDMGEXP" ,"CROPDMG"    ,"CROPDMGEXP" )]

```
### Clean and filter dataset 
Fix datatypes and filter for where fatalities or injuries >0
```{r}
stormDataSet <- stormDataSet[EVTYPE != "?"]
stormDataSet$CROPDMGEXP <- as.character(stormDataSet$CROPDMGEXP)
stormDataSet <- stormDataSet[ INJURIES > 0 | FATALITIES > 0 | PROPDMG > 0 | CROPDMG > 0]

cols <- c("PROPDMGEXP", "CROPDMGEXP")
stormDataSet[,  (cols) := c(lapply(.SD, toupper)), .SDcols = cols]

#convert exponents to numbers

propertyDamageKey <-  c("\"\"" = 10^0,
                 "-" = 10^0, 
                 "+" = 10^0,
                 "0" = 10^0,
                 "1" = 10^1,
                 "2" = 10^2,
                 "3" = 10^3,
                 "4" = 10^4,
                 "5" = 10^5,
                 "6" = 10^6,
                 "7" = 10^7,
                 "8" = 10^8,
                 "9" = 10^9,
                 "H" = 10^2,
                 "K" = 10^3,
                 "M" = 10^6,
                 "B" = 10^9)


cropDamageKey <-  c("\"\"" = 10^0,
                "?" = 10^0, 
                "0" = 10^0,
                "K" = 10^3,
                "M" = 10^6,
                "B" = 10^9)

stormDataSet[, PROPDMGEXP := propertyDamageKey[as.character(stormDataSet[,PROPDMGEXP])]]
stormDataSet[is.na(PROPDMGEXP), PROPDMGEXP := 10^0 ]

stormDataSet[, CROPDMGEXP := cropDamageKey[as.character(stormDataSet[,CROPDMGEXP])] ]
stormDataSet[is.na(CROPDMGEXP), CROPDMGEXP := 10^0 ]

head(stormDataSet)
```
### Calculation costs for crop and property
```{r}

stormDataSet$ECONOMIC_LOSS <- (stormDataSet$CROPDMG*stormDataSet$CROPDMGEXP +     stormDataSet$PROPDMG*stormDataSet$PROPDMGEXP)

stormDataSet$HEALTH_LOSS <- stormDataSet$FATALITIES + stormDataSet$INJURIES
head(stormDataSet)

```

## Events that are Most Harmful to Population Health 
### Summarize data for health impact

```{r}
SummaryHealthLoss <- aggregate(HEALTH_LOSS~EVTYPE, stormDataSet, sum)
SummaryHealthLoss <- SummaryHealthLoss[ order( SummaryHealthLoss$HEALTH_LOSS, decreasing = T),]
head(SummaryHealthLoss, 10)
```

## Results

```{r}
pal <- colorRampPalette(c("red", "#f99494"))
topN <- 10
healthBarChart <- ggplot(SummaryHealthLoss[1:topN,], aes(x=reorder(EVTYPE, -HEALTH_LOSS), y=HEALTH_LOSS))
healthBarChart = healthBarChart + geom_bar(stat="identity", fill=pal(topN), position="dodge")
# y axis
healthBarChart = healthBarChart + ylab("Frequency Count") 
# x-axis 
healthBarChart = healthBarChart + xlab("Event Type") 

healthBarChart = healthBarChart + theme(axis.text.x = element_text(angle=45, hjust=1))
healthBarChart = healthBarChart + ggtitle("Top 10 US Killers") + theme(plot.title = element_text(hjust = 0.5))

print(healthBarChart)

```

## Events that have the Greatest Economic Impact 
### Summarize data for economic impact
```{r}
SummaryEconomicLoss <- aggregate(ECONOMIC_LOSS~EVTYPE, stormDataSet, sum)
SummaryEconomicLoss <- SummaryEconomicLoss[ order( SummaryEconomicLoss$ECONOMIC_LOSS, decreasing = T),]
head(SummaryEconomicLoss, 10)
```

## Results
```{r}
pal <- colorRampPalette(c("red", "#f99494"))
topN <- 10


EconomicBarChart <- ggplot(SummaryEconomicLoss[1:topN,], aes(x=reorder(EVTYPE, -ECONOMIC_LOSS), y=ECONOMIC_LOSS))
EconomicBarChart = EconomicBarChart + geom_bar(stat="identity", fill=pal(topN), position="dodge")
# y axis
EconomicBarChart = EconomicBarChart + ylab("Frequency Count") 
# x-axis 
EconomicBarChart = EconomicBarChart + xlab("Event Type") 

EconomicBarChart = EconomicBarChart + theme(axis.text.x = element_text(angle=45, hjust=1))
EconomicBarChart = EconomicBarChart + ggtitle("Top 10 US Events with Economic impact") + theme(plot.title = element_text(hjust = 0.5))

print(EconomicBarChart)

```